Synthesis of Linear Functional Programs

### 2021-07-18 article/article.tex
Overview first iteration

% General motivation?
Program synthesis is an automated or semi-automated process of deriving a
program, i.e.~generating code, from a high-level specification. Synthesis can be
seen as a means to improve programmer productivity and program correctness
(e.g. through suggestion and autocompletion).
%
Specifications can take many forms such as natural
language~\cite{chen2021evaluating},
examples~\cite{DBLP:conf/popl/FrankleOWZ16} or rich types such as
polymorphic refinement types~\cite{DBLP:conf/pldi/PolikarpovaKS16} or
graded types~\cite{DBLP:conf/lopstr/HughesO20}. Regardless of the kind
of specification, program synthesizers have to deal with two main
inherent sources of complexity -- searching over a vast space of
(potentially) valid programs, and interpreting user intent.

@@

\paragraph{Contributions.}
Despite their long-known potential (\emph{linear types can change the
world!}\cite{Wadler90lineartypes}, \cite{DBLP:journals/mscs/CairesPT16,Bernardy_2018}) and,
primarily, strong proof-theoretic
foundations \cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00}, synthesis of linear functional programs
from linear types in an advanced context has been greatly overlooked.
In this paper we present a framework for synthesis of linear types extended with
recursive algebraic data types, parametric polymorphism and refinements.
% TODO: extensible? proving feasability of... ? showing potential for ... ?

\paragraph{Outline.}
First, we discuss the use of linear types as specifications, the outline of the
synthesis process, and what that entails for the synthetiser -- ilustrated with
concrete results (Section~\ref{sec:overview}). Then, we present the
algorithm and formal system driving the synthesis
(Section~\ref{sec:formal_system}). Aftwerwards, we describe the architecture 
of our synthetiser named \synname, and examine technical details and challenges of the
implementation (Section~\ref{sec:architecture}). Finally, we evaluate our work
through expressiveness benchmarks (i.e.~can we synth $X$?) and time benchmarks
(i.e.~how fast can we synth a term of type $T$?) (Section~\ref{sec:evaluation}).
%TODO: And appendice with...

@@

The core of the synthesis is a \emph{sound} and \emph{complete} system
consisting of \emph{bottom-up} proof search in propositional linear
logic based on \emph{focusing}~\cite{}. Our approach, being grounded
by propositions-as-types, ensures that all synthesized programs
(i.e.~proofs) are well-typed \emph{by construction} (i.e.~if the
synthesis procedure terminates, then the program satisfies its
specification). Moreover, we can leverage the modularity of the
proof-search based approach in two key axes: first, since proof search
need not construct only closed proofs, we can effectively synthesize
program sub-expressions (i.e.~synthesis based on typed holes);
secondly, the framework is amenable to extension beyond the core
propositional language, employing a richer type structure that
includes polymorphism, algebraic data types and type refinements.

\paragraph{A core linear calculus.}
we can synthetise programs from specifications that use
canonical linear types ($\lolli, \with, \tensor, 1, \oplus, \bang$), such as
fundamental linear logic equivalences. For example, given the specification $(A
\lolli (B \lolli C)) \lolli (A \tensor B) \lolli C$, the program
\mintinline{haskell}{\a -> \b -> let c*d = b in a c d}
% $\lambda a .\ \lambda b .$ let $c \tensor d$ = $b$ in $a$ $c$ $d$
(commonly known as \emph{uncurry}) is synthetised.

\paragraph{Beyond propositional logic.}
However, a programming language only with the canonical linear types lacks
expressiveness. To enhance \synname\ and its language, synthetise more
interesting programs -- and thus empirically prove the feasability of more
relevant synthesis in a linear context, we extend the syntax and type system
with recursiveness and richer types: -- polimorfic, refinements, adts
parametrizáveis...  todo

The \synname\ programming language is naturally born alongside the synthetiser
to create an environment for it. Synth goals are inserted in the program by use
of a \emph{synth} keyword or syntatic \emph{mark}, indicating a program should
be generated for a given type. Additionally, it opens the possibility of synthesis with
context, that is, with knowledge of other functions and structures defined in
the same program.

Linear types provide a more certain specification, when compared to simple
traditional types. By requiring some types to be used linearly (consumed) we can
assure, for example, the deconstruction of an ADT. Given the goal $\bang(a \lolli b)
\lolli List\ a \lolli List\ b$, the program \mintinline{haskell}{\x -> \y ->
Nil} wouldn't be valid. To synthetise a valid program,
\mintinline{haskell}{List a}
would have to be deconstructed, and its constructor arguments used, resulting
in the map function without relying on any additional information:

\begin{minted}{haskell}
map c d = let !e = c in
    case d of
          Nil  -> Nil
        | Cons f ->
            let g*h = f in Cons (e g, map (!e) h);
\end{minted}

@@

% \todo[inline]{Encurtar titulo, mas algo nesta onda que descreve
%   ``tudo''}

@@

\mypara{Resourceful Program Synthesis from Graded Linear Types} The
work~\cite{DBLP:conf/lopstr/HughesO20} synthesizes programs using an
approach very similar to our own.  The work employs so-called graded
modal types, which is an refinement of pure linear types -- a more
\emph{fine-grained} version, since it allows for quantitative
specification of resource usage, in contrast to ours either
\emph{linear} or \emph{unrestricted} (via the linear logic
exponential) use of assumptions.  Their resource management is more
complex, and so they provide solutions which adapt Hodas and Miller's
resource management
approach~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09}
-- the model we will be using.

@@

We'll define our inference rules for synthesis under the sequent
calculus system, for it simplifies the work to be done in the
synthesizer.

>>>>

% General motivation?
Program synthesis is an automated or semi-automated process of deriving a
program, i.e.~generating code, from a high-level specification. Synthesis can be
seen as a means to improve programmer productivity and program correctness
(e.g. through suggestion and autocompletion).
%
Specifications can take many forms such as natural
language~\cite{chen2021evaluating},
examples~\cite{DBLP:conf/popl/FrankleOWZ16} or rich types such as
polymorphic refinement types~\cite{DBLP:conf/pldi/PolikarpovaKS16} or
graded types~\cite{DBLP:conf/lopstr/HughesO20}. Regardless of the kind
of specification, program synthesisers have to deal with two main
inherent sources of complexity -- searching over a vast space of
(potentially) valid programs, and interpreting user intent.

@@

\paragraph{Contributions.} Despite their long-known potential (\emph{linear
types can change the world!}\cite{Wadler90lineartypes},
\cite{DBLP:journals/mscs/CairesPT16,Bernardy_2018}) and strong proof-theoretic
foundations
\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00},
synthesis with linear types in a more advanced context
has generally been overlooked in the literature.  In this paper we present a framework for synthesis
with linear types extended with recursive algebraic data types, parametric
polymorphism and refinements.
% TODO: extensible? proving feasability of... ? showing potential for ... ?

\paragraph{Outline.}
First, we discuss linear types as specifications, the outline of the
synthesis process, and what that entails for the synthetiser -- ilustrated with
concrete results (Section~\ref{sec:overview}). Then, we present the
algorithm and formal system driving the synthesis
(Section~\ref{sec:formal_system}). Aftwerwards, we describe the architecture 
of our synthetiser named \synname, and examine technical details and challenges of the
implementation (Section~\ref{sec:architecture}). Finally, we evaluate our work
through expressiveness benchmarks (i.e.~can we synth $X$?) and time benchmarks
(i.e.~how fast can we synth a term of type $T$?) (Section~\ref{sec:evaluation}).
%TODO: And appendice with...

@@

Linear types make for more precise specifications than simple traditional types
because information on which resources must necessarily be used is encoded in
the type. Their preciseness also influences the search space: all programs where
a linear resource is used non-linearly (i.e. not \emph{exactly once}) are
ill-typed. With linearity built-in to the synthesis process, usage of a
linear proposition more than once is not even considered, and unused ones are caught
mid-process, making proof-search in linear logic potentially faster than
in classical logic.  % TODO: Reword?

The core of the synthesis is a \emph{sound} and \emph{complete} system
consisting of \emph{bottom-up} proof-search in propositional linear logic based
on \emph{focusing}~\cite{}. Our approach, being grounded by
propositions-as-types, ensures that all synthetised programs (i.e.~proofs) are
well-typed \emph{by construction} (i.e.~if the synthesis procedure terminates,
then the program satisfies its specification). Moreover, we can leverage the
modularity of the proof-search based approach in two key axes: first, since
proof search need not construct only closed proofs, we can effectively
synthetise program sub-expressions (i.e.~synthesis based on typed holes);
secondly, the framework is amenable to extensions to the core propositional
language, allowing for the introduction of a richer type structure while
preserving the correctness of programs by construction.

The \synname\ programming language is naturally born alongside the synthetiser
to create an environment for it. Synth goals are inserted in the program by use
of a \emph{synth} keyword or mark, indicating a program should be generated for a given
type. Additionally, it opens the possibility of synthesis with context, that is,
with knowledge of other functions and structures defined in the same program.

\paragraph{Core linear calculus.} Initially, we synthetise from
specifications that only use literals and canonical linear types ($\lolli, \with, \tensor, 1,
\oplus, \bang$), such as linear logic theorems. Namely,
the specification $(A \lolli B \lolli C) \lolli (A \tensor B) \lolli C$ produces
the program \mintinline{haskell}{(\a -> \b -> let c*d = b in a c d)}
(known as \emph{uncurry}).

\paragraph{Beyond propositional logic.} To enhance \synname\ and its language,
counter the core's lack of expressiveness, synthetise more interesting programs
-- and thus, empirically prove the feasability of more relevant synthesis in a
linear context, we extend the syntax and type system with recursiveness,
parametric polymorphism, algebraic data types (ADTs) and type refinements.

%TODO: Polymorphism ? read todo below

\paragraph{ADTs.} We reiterate the expressiveness of linear types: by requiring some types to be
used linearly (consumed), we can assure, e.g., the deconstruction of an
ADT. As such, the specification $\bang(a \lolli b) \lolli List\ a \lolli
List\ b$ isn't satisfied by the program \mintinline{haskell}{(\x -> \y -> Nil)}.
To synthetise a valid program from it, \mintinline{haskell}{(List a)} has to be
deconstructed, and its constructor arguments used. \synname\ would output the following
function (\emph{map}) without relying on any additional information:
% TODO: Aqui não estou bem a falar de polymorphism mas parece estranho não por
% visto que falo dos outros temas? se calhar faço um à parte?
%
\begin{minted}{haskell}
map c d = let !e = c in
    case d of
          Nil  -> Nil
        | Cons f -> let g*h = f in Cons (e g, map (!e) h);
\end{minted}

\paragraph{Refinements.} Refinements are predicates added to a type which must hold for any
element that inhabits said type. The \synname\ language supports simple
refinements on integers with arithmetic expressions in the predicates. They
assert the robustness of our main framework, i.e., how it is amenable to
significant additions without necessarily interfering with the algorithm.
%
As an example, from the specification $x \{Int\} \lolli y \{Int\} \lolli z
\{Int\} \lolli k \{Int\ \vert\ x + k = y * z\}$, the
program\ \mintinline{haskell}{(fn a b c = add (mult (sub 0 1) a) (mult c b))} is
synthetised.

\paragraph{Configuration.} Specifications can be additionally augmented with three keywords: ``using'', ``depth'' and
``assert''. Through them, we can fine-tune the synthesis process and, respectively,
force certain functions to be present in the synthesis outcome, allow a
\emph{deeper} search in the valid programs space, and require given predicates
to hold true in the program after the synthesis is complete.

Taken directly from the Linear Haskell paper~\cite{Bernardy_2018}, we present a
complex example of synthesis: given the linear type signatures for array
primitives (\emph{newMArray} passes a new mutable array to a function that uses
it linearly, \emph{write} takes a mutable array and writes a pair to it,
\emph{freeze} consumes a mutable array and produces an imutable array,
\emph{foldl} has the default definition) we try to synthetise a function
``array'' that provides an immutable array given a list of pairs to write. The input program is formulated as
follows:
%
\begin{minted}{haskell}
foldl :: ((a -o b -o a) -> a -o (List b) -o a);
newMArray :: (Int -> (MArray a -o !b) -o b);
write :: (MArray a -o (Int * a) -> MArray a);
freeze :: (MArray a -o !(Array a));
synth array :: (Int -> List (!(Int * a)) -> Array a)
    | using (foldl) | depth 3;
\end{minted}
%
And, in a small fraction of a second, matching exactly the linear definition for
\emph{array} as seen in Linear Haskell~\cite{Bernardy_2018}, \synname\ outputs:
%
\begin{minted}{haskell}
array b c = let !d = b in let !e = c in
    newMArray (!d) (\j -> freeze (foldl (!write) j e));
\end{minted}

@@

@@

\mypara{Resourceful Program Synthesis from Graded Linear Types} The
work~\cite{DBLP:conf/lopstr/HughesO20} synthetises programs using an
approach very similar to our own.  The work employs so-called graded
modal types, which is an refinement of pure linear types -- a more
\emph{fine-grained} version, since it allows for quantitative
specification of resource usage, in contrast to ours either
\emph{linear} or \emph{unrestricted} (via the linear logic
exponential) use of assumptions.  Their resource management is more
complex, and so they provide solutions which adapt Hodas and Miller's
resource management
approach~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09}
-- the model we will be using.

@@

We'll define our inference rules for synthesis under the sequent
calculus system, for it simplifies the work to be done in the
synthetiser.
